Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-20 paper.qmd
Add intro to discussion

The MSE for different binwidths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with default (hyper-)parameter setting (@fig-mnist-mse) were calculated. It is found that tSNE (@fig-mnist-mse (f)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap.

@@

This study makes several important contributions to the field of NLDR. We have developed an algorithm to evaluate the most reasonable NLDR method and (hyper-)parameter choices for creating an accurate \gD{} layout of high-dimensional data. Our objective is to fit a model for the \gD{} layout as a \gD{} wireframe and lifts it into \pD{}, which can be overlaid on the data and visualized using a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

@@

To extend layouts beyond $k$-D, when 2D is clearly inadequate, can use k-means instead of bin centroids.

>>>>

The MSE for different binwidths ($a_1$) using UMAP, PHATE, PaCMAP, TriMAP with default (hyper-)parameter setting and tSNE with default (hyper-)parameter setting and perplexity $89$ (@fig-mnist-mse) were calculated. It is found that tSNE with perplexity $89$ (@fig-mnist-mse (f)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap.

@@

We have developed an algorithm to assess the NLDR method and (hyper-)parameter choice(s) and choose the most accurate representation of the structure(s) present in the \pD{} data. Starting from \gD{} layout, the fitted model, as represented by the positions of points in \kD{}, and turn it into a \pD{} wireframe to overlay on the data, viewing it with a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

@@

To extend layouts beyond \kD{}, when \gD{} is clearly inadequate, can use k-means instead of bin centroids.
